dyngen: a multi-modal simulator for spearheading new single-cell omics analyses

### 2021-05-12 manuscript/supplementary_files.Rmd
rework manuscript

---
title: "dyngen: a multi-modal simulator for spearheading new single-cell omics analyses"
author:
- Robrecht Cannoodt*
- Wouter Saelens*
- Louise Deconinck
- Yvan Saeys
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    citation_package: biblatex
biblatexoptions: sorting=none,url=false
classoption:
- table
- 10pt
- a4paper
header-includes: |
  \usepackage{tcolorbox}
  \usepackage{colortbl}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{fontspec}
  \usepackage{pifont}
  \usepackage{float}
  \usepackage{caption}
  \captionsetup[table]{position=bottom}
  \setmainfont [Path = fonts/,
    UprightFont = *-300,
    ItalicFont = *-300-Italic,
    BoldFont = *-700,
    BoldItalicFont = *-700-Italic
  ]{MuseoSans}

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, results="hold")
library(tidyverse)
```

# Abstract
<!-- format: max. 70 words -->
We present dyngen, a novel, multi-modal simulation engine for studying dynamic cellular processes at single-cell resolution. dyngen is more flexible than current single-cell simulation engines, and allows better method development and benchmarking, thereby stimulating development and testing of novel computational methods. We demonstrate its potential for spearheading novel computational methods on three novel applications: aligning cell developmental trajectories, cell-specific regulatory network inference and estimation of RNA velocity.

# Main text
<!-- format: 1500 Words, no headings, max. 2 figures, max. 20 references-->
Single-cell simulation engines are becoming increasingly important for testing and benchmarking computational methods, a pressing need in the widely expanding field of single-cell biology. Complementary to real biological data, synthetic data provides a valuable alternative where the actual ground truth is completely known and thus can be compared to, in order to make quantitative evaluations of computational methods that aim to reconstruct this ground truth [@zappia_splattersimulationsinglecell_2017]. In addition, simulation engines are more flexible when it comes to stress-testing computational methods, for example by varying the parameters of the simulation, such as the amount of noise, samples, and cells measured, allowing benchmarking of methods over a wide range of possible scenarios. In this way, they can even guide the design of real biological experiments, finding out the best conditions to be used as input for subsequent computational pipelines.

Another, more experimental use of simulation engines is their important role in spearheading the development of novel computational methods, possibly even before real data is available. In this way, simulation engines can be used to assess the value of novel experimental protocols or treatments. Simulation engines are also increasingly important when it comes to finding alternatives to animal models, for example for drug testing and precision medicine. In such scenarios, cellular simulations can act as digital twins, offering unlimited experimentation _in silico_ [@bjornsson_digitaltwinspersonalize_2019].

Simulating realistic data requires that the underlying biology is recapitulated as best as possible, and in the case of transcriptomics data this typically involves modelling the underlying gene regulatory networks. Simulators of "bulk" microarray or RNA-sequencing profiles simulate biological processes (e.g. transcription, translation) by translating a database of known regulatory interactions into a set of ordinary differential equations (ODE) [@roy_systemgeneratingtranscription_2008; @hache_gengesystematicgeneration_2009; @schaffter_genenetweaversilicobenchmark_2011; @vandenbulcke_syntrengeneratorsynthetic_2006]. These methods have been instrumental in performing benchmarking studies [@prill_rigorousassessmentsystems_2010; @marbach_revealingstrengthsweaknesses_2010; @marbach_wisdomcrowdsrobust_2012]. However, the advent of single-cell omics introduced several new types of analyses (e.g. trajectory inference, RNA velocity, cell-specific network inference) which exploit the higher resolution of single-cell versus bulk omics [@luecken_currentbestpractices_2019]. In addition, the data characteristics of single-cell omics are vastly different from bulk omics, typically having much lower library sizes and a higher dropout rate, but also a high number of profiles [@vallejos_normalizingsinglecellrna_2017]. The low library sizes, in particular, are problematic as ODEs are ill-suited for performing low-molecule simulations [@gillespie_exactstochasticsimulation_1977]. This necessitates the development of new single-cell simulators.

To this end, single-cell omics simulators emulate the technical procedures from single-cell omics protocols. Simulators such as Splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2019] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been widely used to compare single-cell methods [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2019; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifysamplesize_2020] and perform independent benchmarks [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018]. However, by focusing more on simulating the single-cell omics protocol (e.g. RNA capture, amplification, sequencing) and less on the underlying biology (e.g. transcription, splicing, translation), their applicability and reusability is limited towards the specific application for which they were designed (e.g. benchmarking clustering or differential expression methods), and extending these tools to include additional modalities or experimental conditions is challenging.

We introduce dyngen, a method for simulating cellular dynamics at a single-cell, single-transcript resolution (Figure\ \ref{fig:overview}). This problem is tackled in three fully-configurable main steps. First, biological processes are mimicked by translating a gene regulatory network into a set of reactions (regulation, transcription, splicing, translation). Second, individual cells are simulated using Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977], which is designed to work well in low-molecule simulations. Finally, real reference datasets are used to emulate single-cell omics profiling protocols.

Throughout a simulation, dyngen tracks many layers of information, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions. In addition, dyngen can simulate a large variety of dynamic processes (e.g. cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g. batch effects and time-series, perturbation and single-cell knockdown experiments). For these reasons, dyngen can cater to a wide range of benchmarking applications, including trajectory inference, trajectory alignment, and trajectory differential expression (Table\ \ref{tab:comparison}).

\begin{figure}[t!]
    \centering
    \includegraphics[width=\linewidth]{result_files/summary/figure_1_edited.pdf}
    \caption{
      \textbf{Showcase of dyngen functionality.}
      \textbf{A:} Changes in abundance levels are driven strictly by gene regulatory reactions.
      \textbf{B:} The input Gene Regulatory Network (GRN) is defined such that it models a dynamic process of interest.
      \textbf{C:} The reactions define how abundance levels of molecules change at any particular time point.
      \textbf{D:} Firing many reactions can significantly alter the cellular state over time.
      \textbf{E:} dyngen keeps track of the likelihood of a reaction firing during small intervals of time, called the propensity, as well as the actual number of firings.
      \textbf{F:} Similarly, dyngen can also keep track of the regulatory activity of every interaction.
      \textbf{G:} A benchmark of trajectory inference methods has already been performed using the cell state ground-truth \cite{saelens_comparisonsinglecelltrajectory_2019}.
      \textbf{H:} The cell state ground-truth enables evaluating trajectory alignment methods.
      \textbf{I:} The reaction propensity ground-truth enables evaluating RNA velocity methods.
      \textbf{J:} The cellwise regulatory network ground-truth enables evaluating cell-specific gene regulatory network inference methods.
    }
    \label{fig:overview}
\end{figure}

<!-- RESULTS -->
We demonstrate dyngen's broad applicability by evaluating three novel types of computational approaches for which no simulation engines exist yet: cell-specific network inference, trajectory alignment and RNA velocity (Figure\ \ref{fig:applications}). We emphasize that our main aim here is to illustrate the potential of dyngen for these evaluations, rather than performing large-scale benchmarking, which would require assessing many more quantitative and qualitative aspects of each method [@weber_essentialguidelinescomputational_2019].

\begin{figure}[t!]
    \centering
    \includegraphics[width=\linewidth]{result_files/summary/figure_2.pdf}
    \caption{
      \textbf{dyngen provides ground-truth data for a variety of applications (left), which can be used to quantitatively evaluate methods (right).}
      \textbf{A:} Trajectory alignment aligns two trajectories between samples. We evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters based on the area differences between the worst possible alignment and the predicted alignment (Area Between Worst And Prediction, or ABWAP).
      \textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving. We evaluated scVelo and velocyto by comparing these vectors with the known velocity vector (velocity correlation) and with the known direction of the cellular trajectory in a dimensionality reduction (velocity arrow cosine).
      \textbf{C:} Cell-specific network inference (CSNI) predicts the regulatory network of every individual cell. We evaluate each cell-specific regulatory network with typical metrics for network inference: the Area Under the Receiver Operating Characteristics-curve (AUROC) and Area Under the Precision Recall-curve (AUPR). We evaluate three CSNI methods by computing the mean AUROC and AUPR across all cells.
    }
    \label{fig:applications}
\end{figure}

**Trajectory alignment** methods align trajectories from different samples and allow studying the differences between the different trajectories. For example, by comparing the transcriptomic profiles of cells from a diseased patient to a healthy control, it might be possible to detect transcriptomics differences (differential expression) of particular cells along a developmental process, or to detect an early stop of the trajectory of the diseased patient. Currently, trajectory alignment is limited to aligning linear trajectories, though other topologies of a trajectory could be aligned as well. Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] is a method designed for aligning temporal sequences for speech recognition but has since been used to compare gene expression kinetics from many different biological processes [@cacchiarelli_aligningsinglecelldevelopmental_2018; @kanton_organoidsinglecellgenomic_2019; @mcfaline-figueroa_pooledsinglecellgenetic_2019; @alpert_alignmentsinglecelltrajectories_2018]. cellAlign [@alpert_alignmentsinglecelltrajectories_2018] uses DTW to perform trajectory alignment, but also includes interpolation and scaling of the single cell data as a preprocessing step. We evaluate the performance of DTW and cellAlign by simulating 40 datasets, each containing two linear trajectories generated with the same gene regulatory network but with slightly different simulation kinetics. We assess the accuracy of the obtained alignments by comparing the generated alignment path with the worst possible alignment that could be performed (ABWAP). A visual guide for this metric can be found in Figure \ \ref{fig:traj_align}D. Overall, cellAlign performs significantly better than DTW (Figure\ \ref{fig:traj_align}, which is likely due to the interpolation and scaling steps provided by cellAlign, reducing noise in the data and improving the comparability of the trajectories. Note that, in this comparison, only linear trajectory alignment is performed. While dyngen can generate non-linear trajectories (e.g. cyclic or branching), both aligning non-linear trajectories and constructing a quantitative accuracy metric for non-linear trajectory alignment is not trivial and an avenue for future work.

**RNA velocity** methods use the relative ratio between pre-mRNA and mature mRNA reads to predict the rate of increase/decrease of RNA molecule abundance, as this can be used to predict the directionality of single cell differentiation in trajectories [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018]. Already two algorithms are currently available for estimating the RNA velocity vector from spliced and unspliced counts: velocyto [@lamanno_rnavelocitysingle_2018] and scvelo [@bergen_generalizingrnavelocity_2020]. Yet, to date, no quantitative assessment of their accuracy has been performed, mainly due to the difficulty in obtaining real ground-truth data to do so. In contrast, the ground-truth RNA velocity can be easily extracted from a dyngen simulation, as it is possible to store the rate at which mRNA molecules are being transcribed and degraded at any particular point in time. We executed velocyto and scvelo (with 2 different parameter settings, stochastic and dynamical) on 42 datasets with a variety of backbones (including linear, bifurcating, cyclic, disconnected). We evaluated the predictions using two metrics (Figure\ \ref{fig:velocity}), one which directly compares the predicted RNA velocity of each gene with the ground-truth RNA velocity (called the "velocity correlation"), and one which compares the direction of the ground-truth trajectory embedded in a dimensionality reduction with the average RNA velocity of cells in that neighbourhood (called the "velocity arrow cosine"). While both velocyto and scvelo obtained high scores for the velocity arrow cosine metric (overall 25th percentile = 0.606), the velocity correlation is rather low (overall 75th percentile = 0.156). This means that predicting the RNA velocity (i.e. transcription rate minus the decay rate) for any particular gene is challenging, but the combined information is very informative in determining the directionality of cell progression in the trajectory. In terms of velocity correlation, no method performed significantly better than the other, whereas "scvelo stochastic" performed slightly worse than "scvelo dynamical" and velocyto in terms of velocity arrow cosine score.

**Cell-specific network inference** (CSNI) methods predict not only which transcription factors regulate which target genes, but also aim to identify how active each interaction is in each of the cells, since interactions can be turned off and on depending on the cellular state. While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of their performance is until now lacking. This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exist that are large enough so that it can be used as a ground-truth for evaluating CSNI methods. Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present. We used this ground-truth to compare the performance of three CSNI methods (Figure\ \ref{fig:scgrn}): LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019], SSN [@liu_personalizedcharacterizationdiseases_2016] and SCENIC [@aibar_scenicsinglecellregulatory_2017]. For each dataset, we computed the mean AUROC and AUPR scores of the individual cells. Comparing the mean AUROC and AUPR showed that pySCENIC significantly outperforms both LIONESS and SSN, and in turn that LIONESS significantly outperforms SSN. The poor performance of SSN is expected, as its methodology for predicting a cell-specific is simply computing the difference in Pearson correlation values applied to the whole dataset and the whole dataset minus one sample. This strategy performs poorly in large datasets where cell correlations are high, as the removal of one cell will not yield large differences in correlation values and will result in mostly noise. Overall, pySCENIC almost always performs better than LIONESS, except for a few datasets where LIONESS does manage to obtain a higher AUROC score. However, by using a different internal network inference (e.g. GENIE3 [@huynh-thu_inferringregulatorynetworks_2010] or pySCENIC's GRNBoost2 [@moerman_grnboost2arboretoefficient_2019]) could significantly increase the performance obtained by LIONESS.

<!-- DISCUSSION -->
In summary, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
However, the framework is flexible enough to be adaptable to a broad range of applications, including methods that integrate clustering, network inference, and trajectory inference. In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015]. Additionally, one could anticipate technological developments in single-cell multi-omics. In this way, dyngen allows designing and evaluating the performance and robustness of new types of computational analyses before experimental data becomes available, comparing which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis. One major assumption of dyngen is that cells are assumed to be well-mixed and independent from each other. Subdividing a cell into multiple 2D or 3D subvolumes or allowing cells to exchange molecules, respectively, could pave the way to better study key cellular processes such as cell division, intercellular communication, and migration [@smith_spatialstochasticintracellular_2019].

# Data Availability
Source data for box plots in Figures\ \ref{fig:applications}, \ref{fig:traj_align}, \ref{fig:velocity} and \ref{fig:scgrn} are provided with this paper. All code and data required to reproduce the analysis are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript). The datasets generated for the different use cases are available on Zenodo with record number 4637926 (doi: [10.5281/zenodo.4637926](https://doi.org/10.5281/zenodo.4637926)).

# Code Availability
dyngen is available as an open-source R package on CRAN at [cran.r-project.org/package=dyngen](https://cran.r-project.org/package=dyngen). The analyses performed in this manuscript are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).

# Author contributions
* W.S. and R.C. designed the study.
* R.C., W.S., and L.D. performed the experiments and analysed the data.
* R.C. and W.S. implemented the dyngen software package.
* R.C., W.S., L.D., and Y.S. wrote the manuscript.
* Y.S. supervised the project.

\newpage

# Methods {#sec:dyngen-methods}
The workflow to generate _in silico_ single-cell data consists of six main steps (Figure\ \ref{fig:explain_methods}).

## Defining the module network {#sec:dyngen-modules}

One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making. The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested _in vivo_. One driver of bifurcation is mutual antagonism, where two genes strongly repress each other [@rekhtman_directinteractionhematopoietic_1999; @xu_regulationbifurcatingcell_2015], forcing one of the two to become inactive [@graf_forcingcellschange_2009]. Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012]. Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other [@yosef_dynamicregulatorynetwork_2013].

To start a dyngen simulation, the user needs to define a module network. The module network describes how sets of genes regulate each other and is what mainly determines which dynamic processes occur within the simulated cells.

A module network consists of modules connected together by regulatory interactions, which can be either up- or down-regulating. A module may have basal expression, which means genes in this module will be transcribed without the presence of transcription factor molecules. A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen-simcell}). At the end of the dyngen process, cells will not be sampled from the burn phase simulations. Interactions between modules have a strength (which is a positive integer) and an effect (+1 for upregulating, -1 for downregulating).

Several examples of module networks are given in Figure\ \ref{fig:example_backbones_onlymodules}.
A simple chain of modules (where one module upregulates the next) results in a _linear_ process. By having the last module repress the first module, the process becomes _cyclic_. Two modules repressing each other is the basis of a _bifurcating_ process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process. Finally, a _converging_ process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.

Note that these examples represent the bare minimum in terms of the number of modules used. Using longer chains of modules is typically desired. In addition, the fate decisions made in this example of a bifurcation is reversible, meaning cells can be reprogrammed to go down a different differentiation path. If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring.

## Generating the gene regulatory network {#sec:dyngen-grn}
The GRN is generated based on the given module network in four main steps (Figure\ \ref{fig:gen_feature_network}).

**Step 1, sampling the transcription factors (TF).** The TFs are the main drivers of the molecular changes in the simulation. The user provides a backbone and the number of TFs to generate. Each TF is assigned to a module such that each module has at least $x$ parameters (default $x=1$). A TF inherits the 'burn' and 'basal expression' from the module it belongs to.

**Step 2, generating the TF interactions.** Let each TF be regulated according to the interactions in the backbone. These interactions inherit the effect, strength, and independence parameters from the interactions in the backbone. A TF can only be regulated by other TFs or itself.

**Step 3, sampling the target subnetwork.**
A user-defined number of target genes are added to the GRN. Target genes are regulated by a TF or another target gene, but are always downstream of at least one TF. To sample the interactions between target genes, one of the many FANTOM5 \cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled. The currently existing TFs are mapped to regulators in the FANTOM5 GRN. The targets are drawn from the FANTOM5 GRN weighted by their page rank value, to create an induced GRN. For each target, at most $x$ regulators are sampled from the induced FANTOM5 GRN (default $x=5$). The interactions connecting a target gene and its regulators are added to the GRN.

**Step 4, sampling the housekeeping subnetwork.**
Housekeeping genes are completely separate from any TFs or target genes. A user-defined set of housekeeping genes is also sampled from the FANTOM5 GRN. The interactions of the FANTOM5 GRN are first subsampled such that the maximum in-degree of each gene is $x$ (default $x=5$). A random gene is sampled and a breadth-first-search is performed to sample the desired number of housekeeping genes.

## Convert gene regulatory network to a set of reactions {#sec:dyngen-reactions}

@@

Simulating a cell's GRN makes use of a stochastic framework which tracks the abundance levels of molecules over time in a discrete quantity. For every gene $G$, the abundance levels of three molecules are tracked, namely of corresponding pre-mRNAs, mature mRNAs and proteins, which are represented by the terms $\x G$, $\y G$ and $\z G$ respectively. The GRN defines how a reaction affects the abundance levels of molecules and how likely it will occur. Gibson and Bruck [@gibson_probabilisticmodelprokaryotic_2000] provide a good introduction to modelling gene regulation with stochastic frameworks, on which many of the concepts below are based.

For every gene in the GRN a set of reactions are defined, namely transcription, splicing, translation, and degradation. Each reaction consists of a propensity function -- a formula $f(.)$ to calculate the probability $f(.) \times \text{d}t$ of it occurring during a time interval $\text{d}t$ -- and the effect -- how it will affect the current state if triggered.

The effects of each reaction mimic the respective biological processes (Table\ \ref{tab:reaction_def}, middle). Transcription of gene $G$ results in the creation of a single pre-mRNA molecule $\x G$. Splicing turns one pre-mRNA $\x G$ into a mature mRNA $\x G$. Translation uses a mature mRNA $\y G$ to produce a protein $\z G$. Pre-mRNA, mRNA and protein degradation results in the removal of a $\x G$, $\y G$, and $\z G$ molecule, respectively.

The propensity of all reactions except transcription are all linear functions (Table\ \ref{tab:reaction_def}, right) of the abundance level of some molecule multiplied by a per-gene constant (Table\ \ref{tab:reaction_params}). The propensity of transcription of a gene $G$ depends on the abundance levels of its TFs. The per-gene and per-interaction constants are based on the median reported production-rates and half-lives of molecules measured of 5000 mammalian genes [@schwanhausser_globalquantificationmammalian_2011], except that the transcription rate has been amplified by a factor of 10.

@@

The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation [@schilstra_biologicgeneexpression_2008], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$. Let $\proptran(\z 1, \z 2, \ldots, \z N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors. The following subsections explain and define the propensity function when $N=1$, $N=2$, and finally for an arbitrary $N$.

### Propensity of transcription when $N=1$
In the simplest case when $N=1$, the promoter can be in one of two states. In state $S_0$, the promoter is not bound by any transcription factors, and in state $S_1$ the promoter is bound by $H_1$. Each state $S_j$ is linked with a relative activation $\alpha_j$, a number between 0 and 1 representing the activity of the promoter at this particular state. The propensity function is thus equal to the expected value of the activity of the promoter multiplied by the pre-mRNA production rate of $G$.

\begin{align}
  f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} \\
\end{align}

For $N=1$, $P(S_1)$ is equal to the Hill equation, where $k_i$ represents the concentration of $H_i$ at half-occupation and $n_i$ represents the Hill coefficient. Typically, $n_i$ is between [1,10]

\begin{align}
  P(S_1) & = \frac{y_1^{n_1}}{k_1^{n_1} + y_1^{n_1}} \\
             & = \frac{(y_1/k_1)^{n_1}}{1 + (y_1/k_1)^{n_1}}
\end{align}

The Hill equation can be simplified by letting $\nu_i = \left(\frac{y_i}{k_i}\right)^{n_i}$.

\begin{align}
P(S_1) & = \frac{\nu_1}{1 + \nu_1} \label{eqn:hillsimp}
\end{align}

Since $P(S_0) = 1 - P(S_1)$, the activation function is formulated and simplified as follows.

\begin{align}
f(y_1) & = \text{xpr} \cdot \left(\alpha_0 \cdot P(S_0) + \alpha_1 \cdot P(S_1)\right) \\
           & = \text{xpr} \cdot \left(\alpha_0 \cdot \frac{1}{1 + \nu_1} + \alpha_1 \cdot \frac{\nu_1}{1 + \nu_1}\right) \\
           & = \text{xpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1}{1 + \nu_1} \\
\end{align}

### Propensity of transcription when $N=2$

When $N=2$, there are four states $S_j$. The relative activations $\alpha_j$ can be defined such that $H_1$ and $H_2$ are independent (additive) or synergistic (multiplicative). In order to define the propensity of transcription $f(.)$, the Hill equation $P(S_j)$ is extended for two transcription factors.

Let $w_j$ be the numerator of $P(S_j)$, defined as the product of all transcription factors bound in that state:

\begin{align}
w_0 & = 1 \\
w_1 & = \nu_1 \\
w_2 & = \nu_2 \\
w_3 & = \nu_1 \cdot \nu_2
\end{align}

The denominator of $P(S_j)$ is then equal to the sum of all $w_j$. The probability of state $S_j$ is thus defined as:

\begin{align}
    P(S_j) & = \frac{w_j}{\sum_{j=0}^{j < 2^N} w_j} \\
               & = \frac{w_j}{1 + \nu_1 + \nu_2 + \nu_1 \cdot \nu_2} \\
               & = \frac{w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)}
\end{align}

Substituting $P(S_j)$ and $w_j$ into $f(.)$ results in the following equation:

\begin{align}
f(y_1, y_2) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
 & = \text{xpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \\
 & = \text{xpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1 + \alpha_2 \cdot \nu_2 + \alpha_3 \cdot \nu_1 \cdot \nu_2}{(\nu_1 + 1) \cdot (\nu_2 + 1)} \\
\end{align}

### Propensity of transcription for an arbitrary $N$
For an arbitrary $N$, there are $2^N$ states $S_j$. The relative activations $\alpha_j$ can be defined such that $H_1$ and $H_2$ are independent (additive) or synergistic (multiplicative). In order to define the propensity of transcription $f(.)$, the Hill equation $P(S_j)$ is extended for $N$ transcription factors.

Let $w_j$ be the numerator of $P(S_j)$, defined as the product of all transcription factors bound in that state:

\begin{align}
  w_j & = \prod_{i=1}^{i \leq N} (j \text{ mod } i) = 1 \text{ ? } \nu_i \text{ : } 1
\end{align}

The denominator of $P(S_j)$ is then equal to the sum of all $w_j$. The probability of state $S_j$ is thus defined as:

\begin{align}
P(S_j) & = \frac{w_j}{\sum_{j=0}^{j < 2^N} w_j} \\
& = \frac{w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)}
\end{align}

Substituting $P(S_j)$ into $f(.)$ yields:

\begin{align}
f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
& = \text{xpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \label{eqn:prop2n}
\end{align}

### Propensity of transcription for a large $N$
For large values of $N$, computing $f(.)$ is practically infeasible as it requires performing $2^N$ summations. In order to greatly simplify $f(.)$, $\alpha_j$ could be defined as 0 when one of the regulators inhibits transcription and 1 otherwise.

\begin{equation}
\alpha_j = \begin{cases}
 0 & \text{ if } \exists i : j \text{ mod } i = 1 \text{ and } H_i \text{ represses } G \\
 1 & \text{otherwise}
\end{cases} \label{eqn:assalpha}
\end{equation}

Substituting equation \ref{eqn:assalpha} into equation \ref{eqn:prop2n} and defining $R = \{1, 2, \ldots, N\}$ and $R^+ = \{i | H_i \text{ activates } G\}$ yields the simplified propensity function:

\begin{align}
f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \frac{\prod_{i \in R^+} (\nu_i + 1)}{\prod_{i \in R} (\nu_i + 1)}
\end{align}

### Independence, synergism and basal expression
The definition of $\alpha_j$ as in equation \ref{eqn:assalpha} presents two main limitations. Firstly, since $\alpha_0 = 1$, it is impossible to tweak the propensity of transcription when no transcription factors are bound. Secondly, it is not possible to tweak the independence and synergism of multiple regulators.

Let $\text{ba} \in [0,1]$ denote the basal expression strength $G$ (i.e. how much will $G$ be expressed when no transcription factors are bound), and $\text{sy} \in [0,1]$ denote the synergism of regulators $H_i$ of $G$, the transcription propensity becomes:

\begin{align}
f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \frac{\text{ba} - \text{sy}^{|R^+|} + \prod_{i \in R^+} (\nu_i + \text{sy})}{\prod_{i \in R} (\nu_i + 1)}
\end{align}

## Simulate single cells {#sec:dyngen-simcell}
dyngen uses Gillespie's stochastic simulation algorithm (SSA) [@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.

Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered. Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j prop_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $prop_j$ the propensity value of the $j$th reaction for the current state of the simulation.

GillespieSSA2 is an optimised library for performing SSA simulations. The propensity functions are compiled to C++ and SSA approximations can be used which allow triggering many reactions simultaneously at each iteration. The framework also allows storing the abundance levels of molecules only after a specific interval has passed since the previous census. By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information. In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained, as well as specific sub-calculations of the propensity values, such as the regulator activity level $reg_{G,H}$.

## Simulate experiment {#sec:dyngen-experiment}
From the SSA simulation we obtain the abundance levels of all the molecules at every state. We need to replicate technical effects introduced by experimental protocols in order to obtain data that is similar to real data. For this, the cells are sampled from the simulations and molecules are sampled for each of the cells. Gene capture rates and library sizes are empirically derived from real datasets as to match real technical variation.

### Sample cells
In this step, $N$ cells are sampled from the simulations. Two approaches are implemented: sampling from an unsynchronised population of single cells (snapshot) or sampling at multiple time points in a synchronised population (time series).

**Snapshot** The backbone consists of several states linked together by transition edges with length $L_i$, to which the different states in the different simulations have been mapped (Figure\ \ref{fig:sample_cells}A). From each transition, $N_i = N / \frac{L_i}{\sum L_i}$ cells are sampled uniformly, rounded such that $\sum N_i = N$.

**Time series** Assuming that the final time of the simulation is $T$, the interval $[0, T]$ is divided into $k$ equal intervals of width $w$ separated by $k-1$ gaps of width $g$. $N_i = N / k$ cells are sampled uniformly from each interval (Figure\ \ref{fig:sample_cells}B), rounded such that $\sum N_i = N$. By default, $k = 8$ and $g = 0.75$. For usual dyngen simulations, $10 \leq T \leq 20$. For larger values of $T$, $k$ and $g$ should be increased accordingly.

### Sample molecules
Molecules are sampled from the simulation to replicate how molecules are experimentally sampled. A real dataset is downloaded from a repository of single-cell RNA-seq datasets [@cannoodt_singlecellomicsdatasets_2018]. For each _in silico_ cell $i$, draw its library size $ls_i$ from the distribution of transcript counts per cell in the real dataset. The capture rate $cr_j$ of each _in silico_ molecule type $j$ is drawn from $N(1, 0.05)$. Finally, for each cell $i$, draw $ls_i$ molecules from the multinomial distribution with probabilities $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.

## Simulating batch effects {#sec:dyngen-batcheffect}
Simulating batch effects can be performed in multiple ways. One such way is to perform the first two steps of the creation of a dyngen model (defining the module network and generating the GRN). For each desired batch, create a separate model for which random kinetics are generated and perform all subsequent dyngen steps (convert to reactions, simulate gold standard, simulate single cells, simulate experiment). Since each separate model has different underlying kinetics, the combined output will resemble having batch effects.

## Determining the ground-truth trajectory {#sec:dyngen-groundtruth}
To construct the ground-truth trajectory, the user needs to provide the ground-truth state network alongside the initial module network (Figure\ \ref{fig:example_backbones}). Each edge in the state network specifies which modules are allowed to change in expression in transitioning from one state to another. For each edge, a simulation is run using the end state of an upstream branch as the initial expression vector, and only allowing the modules as predefined by the attribute to change.

As an example, consider the cyclic trajectory shown in Figure\ \ref{fig:example_backbones}. State S0 begins with an expression vector of all zero values. To simulate the transition from S0 to S1, regulation of the genes in modules A, B and C are turned on. After a predefined period of time, the end state of this transition is considered the expression vector of state S1. To simulate the transition from S1 to S2, regulation of the genes in modules D and E are turned on, while the regulation of genes in module C is turned off. During this simulation, the expression of genes in modules A, B, D, and E is thus allowed to change. The end state of the simulation is considered the expression vector of state S2.

>>>>

---
output:
  pdf_document:
    keep_tex: no
    latex_engine: xelatex
    citation_package: biblatex
biblatexoptions: sorting=none,url=false
classoption:
- table
- 10pt
- a4paper
header-includes: |
  \usepackage{tcolorbox}
  \usepackage{colortbl}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{fontspec}
  \usepackage{pifont}
  \usepackage{float}
  \usepackage{caption}
  \captionsetup[table]{position=bottom}
  \setmainfont [Path = fonts/,
    UprightFont = *-300,
    ItalicFont = *-300-Italic,
    BoldFont = *-700,
    BoldItalicFont = *-700-Italic
  ]{MuseoSans}

@@

```{r setup, include=FALSE}
 # for lua files regarding author information, see https://stackoverflow.com/a/52919256
knitr::opts_chunk$set(collapse=TRUE, results="hold")
library(tidyverse)
```

@@

@@

# Supplementary Figures
\renewcommand{\figurename}{Supplemental Figure}
\renewcommand{\tablename}{Supplemental Table}
